---
title: "R Stream of Decision Modelling for Health Economic Evaluation Reader"
author: "Nichola Naylor"
date: "18/03/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Course Reader 
Text about the course generally & R resouces:

> Opening R projects
Notes for intro pack:
> Intro into the R project and then some code for how to set working directory (R) 
> array, Vector, matrix, variable difference and 1 sentence etc.
> Base r functions used a lot - e.g. rnorm() , rgamma() etc. 
> direct to DART resources
> pictures and bright colours 
> installing packages (and list of packages needed across the exercises)



## R project and working directory



If you want to set the working directory manual you can use the setwd() function 

Alternatively, if you want to set the working directory automatically to the folder your are currently in, you can run the following code

```{r pressure, echo=TRUE}

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

```

## Installing packages 

Although within the course we try to mostly use base R functions, we will need to use packages for some exercises. 

These packages can be installed using the following code. 

```{r pressure, echo=TRUE}

install.packages(tidyr)
install.packages(dplyr)
install.packages(data.table)
install.packages(ggplot2)
install.packages(reshape2)

```

## Key Base R functions

https://www.rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf 

[text description info]

```{r pressure, echo=TRUE}




```
## Random draws

[text description info]

```{r pressure, echo=TRUE}

### r norm... etc.


```


