---
title: "R Stream of Decision Modelling for Health Economic Evaluation Reader - Advanced Course"
author: "Nichola Naylor & Jack Williams"
date: "09/04/2021"
output: html_document
#  pdf_document: default
#  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Course Reader 
Text about the course generally & R resouces:

> Opening R projects
Notes for intro pack:
> Intro into the R project and then some code for how to set working directory (R) 
> array, Vector, matrix, variable difference and 1 sentence etc.
> Base r functions used a lot - e.g. rnorm() , rgamma() etc. 
> direct to DART resources
> pictures and bright colours 
> installing packages (and list of packages needed across the exercises)



## R project and working directory


#### Opening an R project 

Something about R projects here 


#### Working directory

If you download the course material using the R project, then the working directory will automatically be set to the folder where the project is. 

If you want to manually set the working directory, then you can use the setwd() function, or alternatively, if you want to set the working directory automatically to the folder your are currently in, you can run the following code: 
```{r, eval = FALSE} 
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```



## Installing packages 

Although within the course we try to mostly use base R functions, we will need to use packages for some exercises. 

These packages can be installed using the following code. 

```{r, eval = FALSE} 
install.packages(tidyr)
install.packages(dplyr)
install.packages(data.table)
install.packages(ggplot2)
install.packages(reshape2)
```


## Key Base R functions

We will use many base R functions throughout the course. You can see a simple cheat sheet here:
https://www.rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf 



```{r, echo=TRUE}

mean(c(1,2,5))


```
## Probabilistic modelling and sampling from random draws

In module X we will learn about probabilistic modelling, to account for parameter uncertainty in decision models. 

We show some examples of sampling using the normal distribution, beta distribution, and gamma distribution. We run 8 samples of each, but this could be set to any number. 

```{r, echo=TRUE}

# Normal distribution with mean of 20 and standard deviation of 4 
rnorm(n = 8, mean = 20, sd = 4)

# Beta distribution of probability of 0.05 (1 event out of 20 subjects)
rbeta(n = 8, shape1 = 1, shape2 = 19)

# Gamma gistribution for mean of 1000 and SE of 200 
rgamma(n = 8, shape = 25, scale = 40)

```


